---
title: "RTS Distribution Model"
author: "Heidi Rodenhizer"
date: "`r Sys.Date()`"
output: html_document
---

# Load Packages

```{r}
library(biomod2)
library(sf)
library(terra)
library(tidyterra)
library(viridis)
library(ggnewscale)
library(tidyverse)
library(patchwork)
```

```{r}
theme_set(theme_bw())
```


# Load Data

```{r}
pca = rast('./pca/pca.tif')

# artsv1 = st_read('../ARTS/ARTS_main_dataset/v.1.0.1/ARTS_main_dataset_v.1.0.1.geojson') |>
#   st_make_valid() |>
#   distinct() |>
#   # keep one polygon per RTS only
#   filter(CentroidLat == first(CentroidLat),
#          .by = UID)
artsv3 = st_read(
  '../ARTS/ARTS_main_dataset/v.3.1.0/ARTS_main_dataset_v.3.1.0.geojson'
) |>
  st_make_valid() |>
  distinct() |>
  # keep one polygon per RTS only
  filter(CentroidLat == first(CentroidLat), .by = UID)

world = rnaturalearth::ne_countries(scale = "large", returnclass = "sf")
world = world |>
  st_transform(3413) |>
  st_crop(ext(pca))

arctic_boreal = st_read(
  '../../tundra_and_boreal_domain/tundra_and_boreal.shp'
) |>
  st_make_valid()

permafrost = st_read(
  '../../tundra_and_boreal_domain/tundra_and_boreal_Dissolved_permafrost_only/tundra_and_boreal_Dissolved_permafrost_only.shp'
) |>
  st_make_valid()
```


# Approximate ROI from Positive RTS occurrences

```{r}
# roi_v1 = artsv1 |>
#   filter(TrainClass == 'Positive') |>
#   st_buffer(dist = 100000) |>
#   st_union()
#
# st_write(
#   roi_v1,
#   "./sdm/training_data/rts_train_roi_v.1.0.1.geojson"
# )

roi_v1 = st_read(
  "./sdm/training_data/rts_train_roi_v.1.0.1.geojson"
)

# train_v1 = artsv1 |>
#   st_filter(roi_v1,
#             .predicate = st_within)
#
# st_write(
#   train_v1,
#   "./sdm/training_data/rts_train_v.1.0.1.geojson"
# )

train_v1 = st_read(
  "./sdm/training_data/rts_train_v.1.0.1.geojson"
)

# roi_v3 = artsv3 |>
#   filter(TrainClass == 'Positive') |>
#   st_buffer(dist = 100000) |>
#   st_union()

roi_v3 = st_read(
  "./sdm/training_data/rts_train_roi_v.3.1.0.geojson"
)

# st_write(
#   roi_v3,
#   "./sdm/training_data/rts_train_roi_v.3.1.0.geojson"
# )

roi_v3_v1_diff = roi_v1 |>
  mutate(version = "Both") |>
  bind_rows(
    roi_v3 |>
      st_difference(roi_v1) |>
      mutate(version = 'v3 Only')
  )

# train_v3 = artsv3 |>
#   st_filter(roi_v3, .predicate = st_within)

# st_write(
#   train_v3,
#   "./sdm/training_data/rts_train_v.3.1.0.geojson"
# )

train_v3 = st_read(
  "./sdm/training_data/rts_train_v.3.1.0.geojson"
)

# new_rts = train_v3 |>
#   anti_join(
#     train_v1 |>
#       st_drop_geometry()
#   )

# st_write(
#   new_rts,
#   "./sdm/training_data/rts_train_new_v.3.1.0-v.1.0.1.geojson"
# )

new_rts = st_read(
  "./sdm/training_data/rts_train_new_v.3.1.0-v.1.0.1.geojson"
)

world_training_region = world |>
  st_crop(
    roi_v3 |>
      st_buffer(dist = 1000000, joinStyle = "MITRE") |>
      st_bbox()
  )
```

```{r}
artsv1_train_fig = ggplot() +
  geom_sf(data = world_training_region, fill = "gray95") +
  geom_sf(
    data = roi_v1,
    linewidth = 0.2,
    color = 'black',
    fill = 'gray80',
    alpha = 0.5
  ) +
  geom_sf(
    data = train_v1,
    aes(
      color = TrainClass,
      fill = TrainClass
    ),
    size = 0.2,
    linewidth = 0.1
  ) +
  scale_color_manual(
    name = "RTS Presence",
    values = c('cadetblue1', 'red4')
  ) +
  scale_fill_manual(
    name = "RTS Presence",
    values = c('cadetblue1', 'red4')
  ) +
  coord_sf(expand = FALSE) +
  theme(
    legend.position = "none"
  ) +
  facet_wrap(~"ARTS v1")

artsv3_train_fig = ggplot() +
  geom_sf(data = world_training_region, fill = "gray95") +
  geom_sf(
    data = roi_v3,
    linewidth = 0.2,
    color = 'black',
    fill = 'gray80',
    alpha = 0.5
  ) +
  geom_sf(
    data = train_v3,
    aes(
      color = TrainClass,
      fill = TrainClass
    ),
    size = 0.2,
    linewidth = 0.1
  ) +
  scale_color_manual(
    name = "RTS Presence",
    values = c('cadetblue1', 'red4')
  ) +
  scale_fill_manual(
    name = "RTS Presence",
    values = c('cadetblue1', 'red4')
  ) +
  coord_sf(expand = FALSE) +
  theme(
    legend.position = "none"
  ) +
  facet_wrap(~"ARTS v3")

new_rts_train_fig = ggplot() +
  geom_sf(data = world_training_region, fill = "gray95") +
  geom_sf(
    data = roi_v3_v1_diff,
    linewidth = 0.2,
    aes(
      fill = version
    ),
    color = "black",
    alpha = 0.5
  ) +
  scale_fill_manual(
    name = "Training Region\nDifference",
    values = c("Both" = "gray80", "v3 Only" = "gold")
  ) +
  new_scale_fill() +
  geom_sf(
    data = new_rts,
    aes(
      color = TrainClass,
      fill = TrainClass
    ),
    size = 0.2,
    linewidth = 0.1
  ) +
  scale_color_manual(
    name = "RTS Presence",
    values = c('cadetblue1', 'red4')
  ) +
  scale_fill_manual(
    name = "RTS Presence",
    values = c('cadetblue1', 'red4')
  ) +
  coord_sf(expand = FALSE) +
  facet_wrap(~"Added Training Data")

artsv1_train_fig +
  artsv3_train_fig +
  new_rts_train_fig +
  guide_area() +
  plot_layout(guides = 'collect', ncol = 2)

# ggsave(
#   "./figures/gbm_train_data.jpg",
#   height = 6.5,
#   width = 6.5
# )
# ggsave(
#   "./figures/gbm_train_data.pdf",
#   height = 6.5,
#   width = 6.5
# )
```

```{r}
# train_v1 |>
#   count(.by = TrainClass)
# train_v3 |>
#   count(.by = TrainClass)
```


# Aggregate RTS Delineations to Presence/Absence in 1 km Grid

Presence Data

```{r}
# arts_poly_rast = rasterizeGeom(
#   vect(train_v3 %>%
#          filter(
#            TrainClass == 'Positive' &
#              (st_geometry_type(.) == 'POLYGON' | st_geometry_type(.) == 'MULTIPOLYGON'))
#   ),
#   pca,
#   fun = 'count'
# )
#
# arts_point_rast = rasterizeGeom(
#   vect(train_v3 %>%
#          filter(
#            TrainClass == 'Positive' &
#              st_geometry_type(.) == 'POINT')
#   ),
#   pca,
#   fun = 'count'
# )
#
# arts_rast = arts_poly_rast + arts_point_rast
#
# arts_rast = arts_rast |>
#   classify(matrix(c(0, NA), ncol = 2))
#
# # Convert to data frame
# arts_grid = arts_rast |>
#   as.data.frame(xy = TRUE, na.rm = TRUE) |>
#   mutate(rts = case_when(count >= 1 ~ 1,
#                            count == 0 ~ 0))
#
# rm(arts_poly_rast, arts_point_rast)
```

Plot
geom_spatraster isn't great for displaying this data, because the summarizing used to make it plot faster makes a lot of the data look like it's not there

```{r}
# ggplot() +
#   geom_spatraster(data = arts_rast) +
#   geom_sf(data = arts,
#           linewidth = 0.25,
#           size = 0.25,
#           color = 'black',
#           fill = 'transparent') +
#   scale_fill_viridis(na.value = 'transparent') +
#   coord_sf(expand = FALSE)
#
# ggplot() +
#   geom_spatraster(
#     data = arts_rast |>
#       crop(ext(-800000, -650000, -900000, -750000))
#     ) +
#   geom_sf(
#     data = arts |>
#       st_crop(
#         st_bbox(
#           c(
#             xmin = -800000,
#             xmax = -650000,
#             ymin = -900000,
#             ymax = -750000
#           )
#         )
#       ),
#     linewidth = 0.25,
#     size = 0.25,
#     color = 'black',
#     fill = 'transparent'
#     ) +
#   scale_fill_viridis(na.value = 'transparent') +
#   coord_sf(
#     # xlim = c(-800000, -650000),
#     # ylim = c(-900000, -750000),
#     expand = FALSE
#     )
```

Absence Data

```{r}
# arts_neg_rast = rasterize(
#   vect(train_v3 %>%
#          filter(TrainClass == 'Negative') %>%
#          st_centroid()),
#   pca
# )
#
# arts_neg_rast = arts_neg_rast |>
#   classify(matrix(c(1, 0), ncol = 2, byrow = TRUE))
#
# # Make sure that negative 1 km cells don't have any observed RTS in them outside of the negative bounding box
# arts_neg_rast[arts_neg_rast == 0 & !is.na(arts_rast)] = NA
#
# # Convert to data frame
# arts_neg_grid = arts_neg_rast |>
#   as.data.frame(xy = TRUE, na.rm = TRUE) |>
#   rename(count = last) |>
#   mutate(rts = count)
```

Combine Presence and Absence Data

```{r}
# arts_grid = arts_grid |>
#   bind_rows(arts_neg_grid)
#
# write_csv(
#   arts_grid,
#   './sdm/arts_grid_filter_v.3.1.0.csv'
# )
# arts_grid = read_csv('./sdm/arts_grid_filter_v.3.1.0.csv')
```

# Format Data for SDM

```{r}
# arts_biomod = BIOMOD_FormatingData(
#   resp.var = arts_grid |> pull(rts),
#   expl.var = pca |> subset(1:6),
#   resp.xy = arts_grid |> select(x, y),
#   resp.name = 'rts'
# )
#
# saveRDS(arts_biomod, './sdm/arts_biomod_v.3.1.0_filter.rds')
arts_biomod = readRDS('./sdm/arts_biomod_v.3.1.0_filter.rds')
```

# Run Models

```{r}
# print(Sys.time())
# opt.d = bm_ModelingOptions(
#   data.type = 'binary',
#   models = "GBM",
#   strategy = 'default'
# )
#
# print(Sys.time())
# cv_kfold = bm_CrossValidation(
#   arts_biomod,
#   strategy = 'kfold',
#   nb.rep = 1,
#   k = 10,
#   do.full.models = TRUE
# )
#
# print(Sys.time())
# tuned.gbm = bm_Tuning(
#   model = "GBM",
#   tuning.fun = "gbm", ## see in ModelsTable
#   bm.options = opt.d@options$GBM.binary.gbm.gbm,
#   do.formula = FALSE,
#   bm.format = arts_biomod,
#   calib.lines = cv_kfold
# )
#
# print(Sys.time())
#
# saveRDS(tuned.gbm, './sdm/tuned_parameters_v.3.1.0_filter.rds')
```

```{r}
# print(Sys.time())
# opt.t = bm_ModelingOptions(
#   data.type = 'binary',
#   models = "GBM",
#   strategy = "user.defined",
#   user.val = list(GBM.binary.gbm.gbm = tuned.gbm),
#   user.base = "bigboss",
#   bm.format = arts_biomod,
#   calib.lines = cv_kfold
# )
# print(Sys.time())
#
# rts_model= BIOMOD_Modeling(
#   arts_biomod,
#   modeling.id = "GBM 10 K-fold Validation Tuned",
#   models = c("GBM"),
#   CV.strategy = "user.defined",
#   CV.user.table = cv_kfold,
#   OPT.strategy = "user.defined",
#   OPT.user = opt.t,
#   metric.eval = c('TSS','ROC'),
#   var.import = 3,
#   seed.val = 42,
#   nb.cpu = 19
# )
#
# print(Sys.time())
```

```{r}
# saveRDS(rts_model,
#         './sdm/rts_gbm_kfold_10_tuned_v.3.1.0_filter.rds')

rts_model = readRDS('./sdm/rts_gbm_kfold_10_tuned_v.3.1.0_filter.rds')
```

Evaluation Scores

```{r}
model_eval = tibble(rts_model@models.evaluation@val)
model_eval_summary = model_eval |>
  filter(full.name != 'rts_allData_allRun_GBM') |>
  summarise(
    across(
      cutoff:validation,
      list(mean = ~ mean(.x, na.rm = TRUE), sd = ~ sd(.x, na.rm = TRUE))
    ),
    .by = c(metric.eval)
  ) |>
  pivot_longer(cols = cutoff_mean:validation_sd) |>
  separate_wider_delim(name, delim = '_', names = c('variable', 'metric')) |>
  mutate(
    variable = factor(
      variable,
      levels = c(
        'calibration',
        'validation',
        'cutoff',
        'sensitivity',
        'specificity'
      )
    )
  ) |>
  arrange(variable, metric.eval) |>
  pivot_wider(
    id_cols = 'metric',
    names_from = c(variable, metric.eval),
    values_from = value
  )
model_eval_summary
```

Variable Importance

```{r}
var_import = get_variables_importance(rts_model)
var_import_summary = var_import %>%
  filter(full.name == 'rts_allData_allRun_GBM') %>%
  summarise(
    across(var.imp, list(mean = mean, sd = sd), na.rm = TRUE),
    .by = expl.var
  ) %>%
  arrange(var.imp_mean) %>%
  mutate(variable = factor(expl.var, levels = .$expl.var))

var_import_fig = ggplot(var_import_summary, aes(y = variable)) +
  geom_col(aes(x = var.imp_mean)) +
  geom_errorbarh(
    aes(
      xmin = var.imp_mean - var.imp_sd,
      xmax = var.imp_mean + var.imp_sd
    ),
    height = 0.25
  ) +
  scale_x_continuous(
    name = 'Variable Importance',
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_y_discrete(name = 'Variable')

var_import_fig
# ggsave('./figures/variable_importance_filter_v.3.1.0.jpg',
#        var_import_fig,
#        height = 3,
#        width = 4)
# ggsave('./figures/variable_importance_filter_v.3.1.0.pdf',
#        var_import_fig,
#        height = 3,
#        width = 4)

# Represent evaluation scores & variables importance
# bm_PlotEvalMean(bm.out = rts_model)
# bm_PlotEvalBoxplot(bm.out = rts_model, group.by = c('algo', 'algo'))
# bm_PlotEvalBoxplot(bm.out = rts_model, group.by = c('algo', 'run'))
# bm_PlotVarImpBoxplot(bm.out = rts_model, group.by = c('expl.var', 'algo', 'algo'))
# bm_PlotVarImpBoxplot(bm.out = rts_model, group.by = c('expl.var', 'algo', 'run'))
# bm_PlotVarImpBoxplot(bm.out = rts_model, group.by = c('algo', 'expl.var', 'run'))
```

Response Curves

```{r}
response_data = bm_PlotResponseCurves(
  bm.out = rts_model,
  models.chosen = get_built_models(rts_model)[c(1:11)],
  fixed.var = 'median'
)
data_density = response_data[['plot']][['layers']][[2]][['data']]
response_data = response_data[['tab']]

response_data = response_data |>
  mutate(
    run = factor(
      str_split_i(pred.name, '_', 3),
      levels = c(paste0('RUN', seq(1, 10)), 'allRun')
    )
  ) |>
  arrange(run)

response_curve_fig = ggplot(
  response_data,
  aes(x = expl.val, y = pred.val, color = run, linewidth = run)
) +
  geom_line() +
  scale_color_manual(values = c(rep('gray', 10), 'black')) +
  scale_linewidth_manual(values = c(rep(0.2, 10), 0.5)) +
  geom_rug(
    data = data_density,
    aes(x = expl.val),
    linewidth = 0.2,
    sides = 'b',
    inherit.aes = FALSE
  ) +
  scale_x_continuous(name = 'PC Score') +
  scale_y_continuous(name = 'Response') +
  facet_wrap(~expl.name, scales = 'free_x') +
  theme(legend.title = element_blank())

response_curve_fig
# ggsave('./figures/response_curves_filter_v.3.1.0.jpg',
#        response_curve_fig,
#        height = 4,
#        width = 6.5)
# ggsave('./figures/response_curves_filter_v.3.1.0.pdf',
#        response_curve_fig,
#        height = 4,
#        width = 6.5)

bm_PlotResponseCurves(
  bm.out = rts_model,
  models.chosen = get_built_models(rts_model)[11],
  fixed.var = 'median',
  do.bivariate = TRUE
)
```

# Predict Across Circumpolar

```{r}
# print(Sys.time())
# rts_pred <- BIOMOD_Projection(
#   bm.mod = rts_model,
#   proj.name = 'Current',
#   new.env = pca %>% select(1:6),
#   models.chosen = 'all',
#   # models.chosen = 'rts_allData_allRun_GBM',
#   metric.binary = 'all',
#   metric.filter = 'all',
#   build.clamping.mask = TRUE
# )
# print(Sys.time())
# # # rts_pred
# # # plot(rts_pred)
# #
# rts_susceptibility = rast('./rts/proj_Current/proj_Current_rts.tif')
```

```{r}
# saveRDS(rts_pred,
#         './sdm/rts_gbm_kfold_10_pred_tuned_v.3.1.0_filter.rds')

rts_pred = readRDS('./sdm/rts_gbm_kfold_10_pred_tuned_v.3.1.0_filter.rds')

# writeRaster(rts_susceptibility,
#             './sdm/gbm_kfold_10_pred_tuned_combined_v.3.1.0_filter.tif')

rts_susceptibility = rast(
  './sdm/gbm_kfold_10_pred_tuned_combined_v.3.1.0_filter.tif'
)
# rts_prob = median(rts_susceptibility[[1:10]]/10)

# writeRaster(rts_prob,
#             './sdm/gbm_kfold_10_pred_tuned_median_v.3.1.0_filter.tif')
rts_prob = rast(
  './sdm/gbm_kfold_10_pred_tuned_median_v.3.1.0_filter.tif'
)
```

```{r}
rts_prob_m = rast(
  '../makopoulou_data/makopoulou_data_download_20240916/GBM/gbm_aver_raster_calc_10_runs.tif'
) |>
  project(rts_prob, method = 'near')
rts_prob_m = rts_prob_m / 10
```

```{r}
ggplot() +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'gray95'
  ) +
  geom_spatraster(data = rts_prob, maxcell = 1e+07) +
  scale_fill_viridis(
    name = 'RTS Probability\n(%)',
    limits = c(0, 100),
    na.value = 'transparent'
  ) +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'transparent'
  ) +
  coord_sf(expand = FALSE)

# ggsave('./figures/rts_probability_median_filter_v.3.1.0.jpg',
#        height = 5,
#        width = 6.5)

ggplot() +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'gray95'
  ) +
  geom_spatraster(
    data = rts_prob,
    maxcell = 1e+07
  ) +
  scale_fill_viridis(
    name = 'RTS Probability\n(%)',
    limits = c(0, 100),
    na.value = 'transparent'
  ) +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'transparent'
  ) +
  geom_sf(
    data = artsv3 |>
      filter(TrainClass == 'Positive'),
    color = 'black',
    fill = 'black',
    size = 0.5,
    linewidth = 1
  ) +
  coord_sf(
    expand = FALSE,
    # xlim = c(-1800000, -1750000),
    # ylim = c(-600000, -550000)
  )

# # bug makes the fill not line up with the outline when saving to jpg
# ggsave('./figures/rts_probability_median_filter_w_arts_v.3.1.0.jpg',
#        height = 5,
#        width = 6.5)
# ggsave('./figures/rts_probability_median_filter_w_arts_v.3.1.0.pdf',
#        height = 5,
#        width = 6.5)

ggplot() +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'gray95'
  ) +
  geom_spatraster(data = rts_prob_m, maxcell = 1e+07) +
  scale_fill_viridis(
    name = 'RTS Probability\n(%)',
    limits = c(0, 100),
    na.value = 'transparent'
  ) +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'transparent'
  ) +
  coord_sf(expand = FALSE)

```

## v.3.1.0 vs. v.1.0.1

```{r}
rts_prob_v1 = rast(
  './sdm/gbm_kfold_10_pred_tuned_median_v.1.0.1_filter.tif'
)

rts_probability_diff_update = rts_prob -
  rts_prob_v1
```

```{r}
(rts_prob_diff_map = ggplot() +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'gray95'
  ) +
  geom_spatraster(data = rts_probability_diff_update, maxcell = 1e+07) +
  scale_fill_gradient2(
    name = expression(atop(Delta ~ ' RTS Probability', '(v3-v1; %)')),
    limits = c(-100, 100),
    low = '#0066CC',
    mid = '#FFFFDD',
    high = '#990000',
    na.value = 'transparent'
  ) +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'transparent'
  ) +
  coord_sf(expand = FALSE))

# ggsave(
#   './figures/rts_probability_diff_filter_v.3.1.0-v.1.0.1.jpg',
#   rts_prob_diff_map,
#   height = 5,
#   width = 6.5
# )

rts_prob_diff_map +
  new_scale_fill() +
  geom_sf(
    data = new_rts,
    aes(
      color = TrainClass,
      fill = TrainClass
    ),
    size = 0.2,
    linewidth = 0.2
  ) +
  scale_color_manual(
    name = "New RTS\nTraining Data",
    values = c('#003399', '#660000')
  ) +
  scale_fill_manual(
    name = "New RTS\nTraining Data",
    values = c('#003399', '#660000')
  ) +
  coord_sf(expand = FALSE)

# ggsave('./figures/rts_probability_diff_filter_v.3.1.0-v.1.0.1_w_new_rts.jpg',
#        height = 5,
#        width = 6.5)
```

# Classify into Susceptibility Zones
Using <=50%, >50-75%, >75-90%, >90-95%, >95% following Makopoulou et al. 2024

```{r}
rts_quantiles = global(
  rts_prob,
  fun = quantile,
  probs = c(0.50, 0.75, 0.90, 0.95),
  na.rm = TRUE
) |>
  as.numeric()

rts_susceptibility_class = rts_prob |>
  classify(
    rcl = matrix(
      c(
        0,
        rts_quantiles[1],
        1,
        rts_quantiles[1],
        rts_quantiles[2],
        2,
        rts_quantiles[2],
        rts_quantiles[3],
        3,
        rts_quantiles[3],
        rts_quantiles[4],
        4,
        rts_quantiles[4],
        Inf,
        5
      ),
      ncol = 3,
      byrow = TRUE
    )
  )

levels(rts_susceptibility_class) = tibble(
  id = seq(1, 5),
  susceptibility_class = c(
    'Very Low (<=50%)',
    'Low (>50-75%)',
    'Medium (>75-90%)',
    'High (>90-95%)',
    'Very High (>95%)'
  )
)

# writeRaster(rts_susceptibility_class,
#             './sdm/gbm_kfold_10_pred_tuned_combined_class_v.3.1.0_filter.tif')
```

```{r, fig.height = 6, fig.width = 6.5}
ggplot() +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'gray95'
  ) +
  geom_spatraster(
    data = rts_susceptibility_class,
    maxcell = 1e+07
  ) +
  scale_fill_viridis(
    name = 'RTS Susceptibility',
    discrete = TRUE,
    na.value = 'transparent',
    na.translate = FALSE
  ) +
  # scale_fill_manual(name = 'RTS Susceptibility',
  #                   values = c('#003388',
  #                              '#3399FF',
  #                              '#FFFF33',
  #                              '#FF6600',
  #                              '#CC0000'),
  #                   na.value = 'transparent',
  #                   na.translate = FALSE) +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'transparent'
  ) +
  coord_sf(expand = FALSE) +
  theme(
    legend.position = 'inside',
    legend.justification = c(1, 0),
    legend.position.inside = c(0.998, 0.002)
  )

# ggsave('./figures/rts_susceptibility_filter_v.3.1.0.jpg',
#        height = 6,
#        width = 6.5)
# ggsave('./figures/rts_susceptibility_filter_blue_red_v.3.1.0.jpg',
#        height = 6,
#        width = 6.5)
```

# Classify into Presence/Absence

Use the mean of the ROC and TSS cut-off values as the cut-off between presence and absence.

```{r}
cutoff = round(mean(c(
  model_eval_summary$cutoff_ROC[1] / 10,
  model_eval_summary$cutoff_TSS[1] / 10
)))

rts_presence_pred = rts_prob |>
  classify(
    rcl = matrix(
      c(0, cutoff, 0, cutoff, 1000, 1),
      byrow = TRUE,
      ncol = 3
    )
  )

levels(rts_presence_pred) = tibble(
  id = seq(0, 1),
  rts_presence = c('No RTS', 'RTS')
)

# Multiple cut-offs for comparison
cutoffs = seq(10, 50, by = 10)

classify_rts = function(prediction, cutoff) {
  rts_presence_pred = prediction |>
    classify(
      rcl = matrix(
        c(0, cutoff, 0, cutoff, 1000, 1),
        byrow = TRUE,
        ncol = 3
      )
    )

  levels(rts_presence_pred) = tibble(
    id = seq(0, 1),
    rts_presence = c('No RTS', 'RTS')
  )

  return(rts_presence_pred)
}

rts_presence_list = map(
  cutoffs,
  ~ classify_rts(rts_prob, .x)
)

# writeRaster(rts_presence_pred,
#             './sdm/rts_presence_pred_filter_v.3.1.0.tif')
```

```{r}
rts_area = map2_df(
  rts_presence_list,
  cutoffs,
  ~ zonal(
    cellSize(.x),
    .x,
    fun = sum
  ) |>
    mutate(
      area = area / (1000 * 1000),
      cutoff = .y
    ) # convert to km^2 from m^2
)

rts_area = rts_area |>
  mutate(
    susceptible_percent_roi = area / sum(area),
    .by = cutoff
  ) |>
  filter(rts_presence == "RTS") |>
  mutate(
    susceptible_percent_arctic_boreal = area /
      as.numeric(
        arctic_boreal |>
          st_transform(crs = 3413) |>
          mutate(area = st_area(arctic_boreal) / (1000 * 1000)) |> # convert to km^2 from m^2
          summarise(area = sum(area)) |>
          pull(area)
      ),
    susceptible_percent_permafrost = area /
      as.numeric(
        permafrost |>
          st_transform(crs = 3413) |>
          mutate(area = st_area(permafrost) / (1000 * 1000)) |> # convert to km^2 from m^2
          summarise(area = sum(area)) |>
          pull(area)
      )
  )

# write_csv(
#   rts_area,
#   "./tables/rts_coverage.csv"
# )
```

```{r}
ggplot() +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'gray95'
  ) +
  geom_spatraster(
    data = rts_presence_pred,
    maxcell = 1e+07
  ) +
  scale_fill_viridis(
    name = 'Predicted RTS Presence',
    discrete = TRUE,
    na.value = 'transparent',
    na.translate = FALSE
  ) +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'transparent'
  ) +
  coord_sf(expand = FALSE) +
  theme(
    legend.position = 'inside',
    legend.justification = c(1, 0),
    legend.position.inside = c(0.998, 0.002)
  )

# ggsave('./figures/rts_presence_pred_filter_v.3.1.0.jpg',
#        height = 6,
#        width = 6.5)
```

# Coefficient of Variation

```{r}
rts_cov = stdev(rts_susceptibility[[1:10]]) / mean(rts_susceptibility[[1:10]])
```

```{r}
ggplot() +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'gray95'
  ) +
  geom_spatraster(
    data = rts_cov,
    maxcell = 1e+07
  ) +
  scale_fill_viridis(
    name = 'Coefficient of Variation',
    na.value = 'transparent'
  ) +
  # scale_fill_gradient(name = 'Coefficient of Variation',
  #                     # limits = c(0, 1.15),
  #                     low = '#FFF5DD',
  #                     high = '#990000',
  #                     na.value = 'transparent') +
  geom_sf(
    data = world,
    aes(geometry = geometry),
    inherit.aes = FALSE,
    fill = 'transparent'
  ) +
  coord_sf(expand = FALSE)

# ggsave('./figures/rts_coefficient_of_variation_filter_v.3.1.0.jpg',
#        height = 5,
#        width = 6.5)

# ggsave('./figures/rts_coefficient_of_variation_filter_red_v.3.1.0.jpg',
#        height = 5,
#        width = 6.5)
```
