---
title: "Analyze RTS Susceptibility Results"
author: "Heidi Rodenhizer"
date: '`r Sys.Date()`'
output: html_document
---

# Load Packages

```{r}
library(terra)
library(tidyterra)
library(sf)
library(viridis)
library(tidyverse)
```

```{r}
theme_set(theme_bw())
```


# Load Data

```{r}
rts_susceptibility = rast('./sdm/gbm_kfold_10_pred_tuned_combined_class_v.1.0.1.tif')
rts_susceptibility_m = rast('../makopoulou_data/Susceptibility_Map.tif') |>
  project(rts_susceptibility, method = 'near')
rts_prob = rast('./sdm/gbm_kfold_10_pred_tuned_combined_v.1.0.1.tif')
rts_prob = median(rts_prob[[1:10]]/10)
rts_prob_m = rast('../makopoulou_data/makopoulou_data_download_20240916/GBM/gbm_aver_raster_calc_10_runs.tif') |>
  project(rts_prob, method = 'near')
rts_prob_m = rts_prob_m/10
```

```{r}
arts = st_read('../ARTS/ARTS_main_dataset/v.1.0.1/ARTS_main_dataset_v.1.0.1.geojson')
pca = rast('./pca/pca.tif')
pca_model = readRDS('./pca/pca.Rdata')
```

```{r}
m_env = c(
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/bd.tif'),
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/fdd.tif'),
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/fine.tif'),
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/rainfall.tif'),
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/slop90_geom_A.tif'),
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/tdd.tif')
)
```

```{r}
files = list.files(
  '../makopoulou_data/RTS_presences_not_in_ARTS',
  full.names = TRUE,
  recursive = TRUE,
  pattern = '\\.shp$'
)
files = files[which(!str_detect(files, 'Couture.+((MVLSD)|(ald))'))]
rts_not_in_arts = map_dfr(
  files,
  ~ .x |>
    st_read() |>
    select(geometry) |>
    mutate(source = str_split_i(.x, '/', i = -1)) |>
    st_transform(crs = 3413)
) |>
  st_as_sf()

mrts = st_read('../makopoulou_data/makopoulou_data_download_20240916/rts_presences/RTSs_abs_pres.shp') |>
  st_transform(crs = 3413)
```

```{r}
world = rnaturalearth::ne_countries(scale = "large", returnclass = "sf")
world = world |>
  st_transform(3413) |>
  st_crop(ext(rts_susceptibility))
```


# Difference in RTS Susceptibility

```{r}
rts_susceptibility_diff = rts_susceptibility_m - rts_susceptibility
rts_prob_diff = rts_prob_m - rts_prob
```

```{r}
# arts_points = arts %>%
#    filter(CentroidLat == first(CentroidLat),
#          .by = UID) %>%
#   st_centroid(.)
# 
# marts_merge_1 = arts_points |>
#   st_join(mrts, join = st_is_within_distance, dist = 10, left = TRUE)
# 
# marts_merge_2 = mrts |>
#   st_join(arts, join = st_is_within_distance, dist = 10, left = TRUE)
# 
# marts_merge = marts_merge_1 |>
#   bind_rows(marts_merge_2) |>
#   distinct() |>
#   mutate(
#     TrainClassMakopoulou = case_when(
#       Bin == 0 ~ 'Negative',
#       Bin == 1 ~ 'Positive',
#       is.na(Bin) ~ NA
#     ),
#     dataset = case_when(
#       !is.na(TrainClass) & !is.na(TrainClassMakopoulou) ~ 'Both',
#       !is.na(TrainClass) & is.na(TrainClassMakopoulou) ~ 'ARTS',
#       is.na(TrainClass) & !is.na(TrainClassMakopoulou) ~ 'Makopoulou'
#     )
#   )
# 
# st_write(marts_merge, './arts_makopoulou_merge/arts_makopoulou_merge.geojson')
marts_merge = st_read('./arts_makopoulou_merge/arts_makopoulou_merge.geojson')
```

## Map Difference

```{r, fig.height = 6, fig.width = 6}
ggplot() +
  geom_sf(data = world) +
  geom_spatraster(
    data = rts_prob_diff,
    maxcell = 1e+07
  ) +
  scale_fill_gradient2(
    name = expression(Delta~' RTS Probability'),
    limits = c(-100, 100),
    low = '#0066CC',
    mid = '#FFFFDD',
    high = '#990000',
    na.value = 'transparent'
  ) +
  coord_sf(expand = FALSE) +
  ggtitle('RTS Probability Difference (Makopoulou - Woodwell)') +
  theme(legend.position = 'inside',
        legend.justification = c(1, 0),
        legend.position.inside = c(0.999, 0.001))
# ggsave('./figures/rts_prob_diff.jpg',
#        height = 6,
#        width = 6)

ggplot() +
  geom_sf(data = world) +
  geom_spatraster(
    data = rts_susceptibility_diff,
    maxcell = 1e+07
  ) +
  scale_fill_gradient2(
    name = expression(Delta~'RTS Susceptibility'),
    limits = c(-5, 5),
    low = '#0066CC',
    mid = '#FFFFDD',
    high = '#990000',
    na.value = 'transparent'
  ) +
  coord_sf(expand = FALSE) +
  ggtitle('RTS Susceptibility Difference (Makopoulou - Woodwell)')
```

```{r}
ggplot() +
  geom_sf(data = world) +
  geom_spatraster(
    data = rts_prob_diff,
    # maxcell = 1e+07
  ) +
  scale_fill_gradient2(
    name = expression(Delta~' RTS Probability'),
    limits = c(-100, 100),
    low = '#0066CC',
    mid = '#FFFFDD',
    high = '#990000',
    na.value = 'transparent'
  ) +
  geom_sf(
    data = marts_merge, 
    aes(color = dataset), 
    size = 0.1
    ) +
  # scale_color_manual(
  #   name = 'RTS Not in ARTS',
  #   values = c('black')
  #   ) +
  coord_sf(expand = FALSE) +
  ggtitle('RTS Probability Difference (Makopoulou - Woodwell)')

# ggsave('./figures/rts_susceptibility_diff.jpg',
#        height = 6,
#        width = 8)
```

## Extract Underlying Variables for Principal Components

```{r}
components = pca_model$rotation |>
  as_tibble() |>
  mutate(
    variable = rownames(pca_model$rotation),
    .before = PC1
    ) |>
  select(variable:PC6) |>
  pivot_longer(PC1:PC6, names_to = 'pc', values_to = 'correlation') |>
  arrange(pc, -abs(correlation))

sd_factor = 1.5
main_components = components |>
  filter(correlation <= -1*sd_factor*sd(correlation) | correlation >= sd_factor*sd(correlation),
         .by = pc) |>
  mutate(order = seq(1:n()),
         label = paste0(variable, ': ', round(correlation, 2)),
         .by = pc) # |>
  # pivot_wider(
  #   id_cols = order, 
  #   names_from = PC, 
  #   values_from = c(variable, correlation),
  #   names_vary = 'slowest',
  #   names_glue = '{PC}_{.value}'
  #   )

# write_csv(
#   components,
#   './tables/pc_variable_correlations.csv'
# )
# write_csv(
#   main_components,
#   './tables/pc_main_components.csv'
# )
```

```{r}
ggplot(components,
       aes(x = correlation)) +
  geom_vline(xintercept = 0,
             color = 'gray50',
             linetype = 'dashed') +
  geom_histogram(bins = 30) +
  geom_histogram(
    data = main_components,
    aes(x = correlation),
    bins = 30,
    inherit.aes = FALSE,
    color = 'transparent',
    fill = 'black'
  ) +
  # geom_text(
  #   data = main_components,
  #   aes(x = correlation, y = 2.5, label = variable),
  #   angle = 90,
  #   hjust = 0,
  #   inherit.aes = FALSE
  #   ) +
  facet_wrap(~ pc, ncol = 2)
```

## Model PC Scores Relative to RTS Probability Difference

```{r}
rts_prob_diff_df = rts_prob_diff |>
  as.data.frame(
  xy = TRUE
) |>
  filter(!is.na(gbm_1_extr)) |>
  rename(prob_diff = gbm_1_extr) |>
  mutate(
    across(x:y,
           ~ round(.x, 8)),
    prob_diff_round = round(prob_diff/10, 0)*10,
    prob_diff_class = factor(prob_diff_round)
    )

pca_df = pca[[1:6]] |>
  mask(rts_prob_diff) |>
  as.data.frame(
    xy = TRUE
  ) |>
  filter(!is.na(PC1)) |>
  mutate(
    across(x:y,
           ~ round(.x, 8))
    )

rts_pca_df = rts_prob_diff_df |>
  left_join(pca_df,
            by = c('x', 'y')) |>
  pivot_longer(PC1:PC6, names_to = 'pc', values_to = 'value')
```

```{r}
rts_pca_df_summary = rts_pca_df |>
  summarize(
    value_mean = mean(value, na.rm = TRUE),
    value_median = median(value, na.rm = TRUE),
    value_sd = sd(value, na.rm = TRUE),
    .by = c(prob_diff_round, prob_diff_class, pc)
  )
```

```{r}
ggplot(
  rts_pca_df,
       aes(x = pc, 
           y = value, 
           color = prob_diff_class, 
           group = interaction(prob_diff_class, pc))
  ) +
  geom_boxplot(position = position_dodge()) +
  scale_y_continuous(name = 'PC Score') +
  scale_color_viridis(name = 'RTS\nProbability\nDifference',
                      discrete = TRUE) +
  theme(axis.title.x = element_blank())
```

```{r}
min_lm = lm(
  value_mean ~ 1,
  data = rts_pca_df_summary
)

max_lm = lm(
  value_mean ~ (prob_diff_round + I(prob_diff_round^2) + I(prob_diff_round^3))*pc,
  data = rts_pca_df_summary
)

rts_suscep_diff_summary_lm = step(
  max_lm,
  list(lower = min_lm, upper = max_lm)
)
summary(rts_suscep_diff_summary_lm)

rts_suscep_diff_lm_params = emmeans::emmeans(
  rts_suscep_diff_summary_lm,
  ~ pc,
  at = list(prob_diff_round = 0)
) |>
  as_tibble() |>
  rename(intercept = emmean) |>
  rename_with(~ paste0('int_', .x),
              .cols = SE:upper.CL) |>
  full_join(
    emmeans::emtrends(
      rts_suscep_diff_summary_lm, 
      ~ pc, 
      var = "prob_diff_round"
    ) |>
      as_tibble() |>
      rename(slope = prob_diff_round.trend) |>
      rename_with(~ paste0('slope_', .x),
                  .cols = SE:upper.CL)
  ) |>
  full_join(
    emmeans::emtrends(
      rts_suscep_diff_summary_lm, 
      ~ pc, 
      var = "I(prob_diff_round^2)"
    ) |>
      as_tibble() |>
      rename(quadratic_term = `I(prob_diff_round^2).trend`) |>
      rename_with(~ paste0('quadratic_', .x),
                  .cols = SE:upper.CL)
  ) |>
  full_join(
    emmeans::emtrends(
      rts_suscep_diff_summary_lm, 
      ~ pc, 
      var = "I(prob_diff_round^3)"
    ) |>
      as_tibble() |>
      rename(cubic_term = `I(prob_diff_round^3).trend`) |>
      rename_with(~ paste0('cubic_', .x),
                  .cols = SE:upper.CL)
  )
```

The Makopoulou model tends to predict higher RTS Susceptibility with low PC1, PC3, and PC4 values and high PC2 and PC5 values. In terms of input variables this means that the Makopoulou model has higher RTS Susceptibility when:
- precip sum, max, MAGT, runoff sum, and ALT are larger (PC1) - Marine influenced areas
- temp seasonality, tmax, srad sum, and LST are lower; and tmin and runoff max are higher (PC2) - Marine influenced areas
- the trend in soil moisture and LST over time are lower (PC5) - Less climate change related increase in precip/temperature
- the trend in precip, precip max, and runoff over time are higher; and silt values are higher (PC3)
- the trend in runoff and tmax over time are higher; silt and MAGT are lower; and soil moisture and said are higher

```{r, fig.height = 6, fig.width = 6.5}
labels = tibble(
      pc = paste0('PC', c(seq(1, 6), 6)),
      x = c(rep(NA, 5), -5, 5),
      hjust = c(rep(NA, 5), 1, 0),
      xend = c(rep(NA, 5), -40, 40),
      y = c(rep(NA, 5), -8, -8),
      label = c(rep(NA, 5), 'Woodwell Greater', 'Makopoulou Greater'),
      yarrow = c(rep(NA, 5), -7, -7),
      )

ggplot(
  rts_pca_df_summary,
       aes(x = prob_diff_round, 
           y = value_mean)
  ) +
  geom_vline(xintercept = 0,
             color = 'gray50',
             linetype = 'dashed') +
  # geom_point(
  #   aes(y = value_median),
  #   size = 2,
  #   color = 'red'
  #   ) +
  geom_point(size = 0.5) +
  geom_errorbar(
    aes(ymin = value_mean - value_sd, ymax = value_mean + value_sd)
    ) +
  geom_smooth(
    method = 'lm', 
    formula = y ~ (x + I(x^2) + I(x^3)),
    color = 'black',
    linewidth = 0.5
    ) +
  # geom_smooth(
  #   method = 'gam', 
  #   linewidth = 0.5
  #   ) +
  geom_text(
    data = labels,
    aes(x = x, y = y, label = label, hjust = hjust),
    size = 2.5
  ) +
  geom_segment(
    data = labels,
    aes(x = x, xend = xend, y = yarrow, yend = yarrow),
    arrow = arrow(length = unit(0.05, 'inches'))
    ) +
  geom_text(
    data = main_components,
    aes(x = -100, y = -3.5 - order*0.75, label = label),
    hjust = 0,
    vjust = 0,
    size = 2.5
  ) +
  scale_x_continuous(
    name = 'RTS Probability Difference\n(Makopoulou-Woodwell)',
    # breaks = seq(-4, 4, by = 2),
    ) +
  scale_y_continuous(name = 'PC Score') +
  facet_wrap(~ pc, ncol = 3) +
  ggtitle('PC Scores by Mean RTS Probability Difference\nWith Cubic Linear Models')

# ggsave('./figures/rts_prob_diff_models.jpg',
#        height = 6,
#        width = 6.5)
```

## Model Environmental Variables Relative to RTS Probability Difference

```{r}
m_env_df = m_env[[1:6]] |>
  project(rts_prob_diff) |>
  mask(rts_prob_diff) |>
  as.data.frame(
    xy = TRUE
  ) |>
  filter(!is.na(bd) & !is.na(fdd) & !is.na(fine) & !is.na(rainfall) & !is.na(slop90_geom_A) & !is.na(tdd)) |>
  mutate(
    across(x:y,
           ~ round(.x, 8)),
    across(bd:tdd, 
           ~ (.x - mean(.x, na.rm = TRUE))/sd(.x))
    )

rts_m_env_df = rts_prob_diff_df |>
  left_join(m_env_df,
            by = c('x', 'y')) |>
  pivot_longer(bd:tdd, names_to = 'variable', values_to = 'value')
```

```{r}
rts_m_env_df_summary = rts_m_env_df |>
  summarize(
    value_mean = mean(value, na.rm = TRUE),
    value_median = median(value, na.rm = TRUE),
    value_sd = sd(value, na.rm = TRUE),
    .by = c(prob_diff_round, prob_diff_class, variable)
  )
```

```{r}
ggplot(
  rts_m_env_df_summary,
       aes(x = prob_diff_round, 
           y = value_mean)
  ) +
  geom_vline(xintercept = 0,
             color = 'gray50',
             linetype = 'dashed') +
  # geom_point(
  #   aes(y = value_median),
  #   size = 2,
  #   color = 'red'
  #   ) +
  geom_point(size = 0.5) +
  geom_errorbar(
    aes(ymin = value_mean - value_sd, ymax = value_mean + value_sd)
    ) +
  geom_smooth(
    method = 'lm', 
    formula = y ~ (x + I(x^2) + I(x^3)),
    color = 'black',
    linewidth = 0.5
    ) +
  # geom_text(
  #   data = labels,
  #   aes(x = x, y = y, label = label, hjust = hjust),
  #   size = 2.5
  # ) +
  # geom_segment(
  #   data = labels,
  #   aes(x = x, xend = xend, y = yarrow, yend = yarrow),
  #   arrow = arrow(length = unit(0.05, 'inches'))
  #   ) +
  # geom_text(
  #   data = main_components,
  #   aes(x = -100, y = -3.5 - order*0.75, label = label),
  #   hjust = 0,
  #   vjust = 0,
  #   size = 2.5
  # ) +
  scale_x_continuous(
    name = 'RTS Probability Difference\n(Makopoulou-Woodwell)',
    # breaks = seq(-4, 4, by = 2),
    ) +
  scale_y_continuous(name = 'Environmental Variable Z-Score') +
  facet_wrap(~ variable, ncol = 3) +
  ggtitle('Environmental Variable Scores by Mean RTS Probability Difference\nWith Cubic Linear Models')

# ggsave('./figures/rts_prob_diff_models_makopoulou.jpg',
#        height = 6,
#        width = 6.5)

```

## Map Top Predictors of RTS in Our Model

```{r}
ggplot() +
  geom_sf(data = world) +
  geom_spatraster_rgb(
    data = pca,
    r = 1,
    g = 2,
    b = 5,
    max_col_value = 7
  ) +
  # geom_sf(data = rts_susceptibility_diff_extremes_poly,
  #         aes(color = susceptibility_class),
  #         fill = 'transparent') +
  coord_sf(expand = FALSE) +
  ggtitle('Principal Components 1, 2, and 5 as RGB')

# ggsave('./figures/top_predictors_rgb.jpg',
#        height = 6.5,
#        width = 6.5)
```

