---
title: "Analyze RTS Susceptibility Results"
author: "Heidi Rodenhizer"
date: '`r Sys.Date()`'
output: html_document
---

# Load Packages

```{r}
library(terra)
library(tidyterra)
library(sf)
library(viridis)
library(tidyverse)
```

```{r}
theme_set(theme_bw())
```


# Load Data

```{r}
rts_susceptibility = rast('./sdm/gbm_kfold_10_pred_tuned_combined_class_v.1.0.1.tif')
rts_susceptibility_m = rast('../makopoulou_data/Susceptibility_Map.tif') |>
  project(rts_susceptibility, method = 'near')
rts_prob = rast('./sdm/gbm_kfold_10_pred_tuned_combined_v.1.0.1.tif')
rts_prob = median(rts_prob[[1:10]]/10)
rts_prob_m = rast('../makopoulou_data/makopoulou_data_download_20240916/GBM/gbm_aver_raster_calc_10_runs.tif') |>
  project(rts_prob, method = 'near')
rts_prob_m = rts_prob_m/10
```

```{r}
arts = st_read(
  '../ARTS/ARTS_main_dataset/v.1.0.1/ARTS_main_dataset_v.1.0.1.geojson'
  ) |>
  st_make_valid()
arts_grid = read_csv('./sdm/arts_grid.csv')
pca = rast('./pca/pca.tif')
pca_model = readRDS('./pca/pca.Rdata')
```

```{r}
m_env = c(
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/bd.tif'),
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/fdd.tif'),
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/fine.tif'),
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/rainfall.tif'),
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/slop90_geom_A.tif'),
  rast('../makopoulou_data/makopoulou_data_download_20240916/rts_var/tdd.tif')
)
```

```{r}
# files = list.files(
#   '../makopoulou_data/RTS_presences_not_in_ARTS',
#   full.names = TRUE,
#   recursive = TRUE,
#   pattern = '\\.shp$'
# )
# files = files[which(!str_detect(files, 'Couture.+((MVLSD)|(ald))'))]
# rts_not_in_arts = map_dfr(
#   files,
#   ~ .x |>
#     st_read() |>
#     select(geometry) |>
#     mutate(source = str_split_i(.x, '/', i = -1)) |>
#     st_transform(crs = 3413)
# ) |>
#   st_as_sf()

mrts = st_read('../makopoulou_data/makopoulou_data_download_20240916/rts_presences/RTSs_abs_pres.shp') |>
  st_transform(crs = 3413)
```

```{r}
world = rnaturalearth::ne_countries(scale = "large", returnclass = "sf")
world = world |>
  st_transform(3413) |>
  st_crop(ext(rts_susceptibility))
```


# Difference in RTS Probability

```{r}
rts_susceptibility_diff = rts_susceptibility_m - rts_susceptibility
rts_prob_diff = rts_prob_m - rts_prob
```

## Prep RTS Points for Comparison

```{r}
# # RTS
# mrts_rast = rasterizeGeom(
#   vect(mrts |>
#          filter(Bin == 1)),
#   pca,
#   fun = 'count'
# )
# 
# mrts_rast = mrts_rast |>
#   classify(matrix(c(0, NA), ncol = 2))
# 
# # Convert to data frame
# mrts_grid = mrts_rast |>
#   as.data.frame(xy = TRUE, na.rm = TRUE) |>
#   rename(count = lyr1) |>
#   mutate(rts = case_when(count >= 1 ~ 1,
#                            count == 0 ~ 0))
# 
# # RTS Absences
# mrts_neg_rast = rasterize(
#   vect(mrts %>%
#          filter(Bin == 0)),
#   pca
# )
# 
# mrts_neg_rast = mrts_neg_rast |>
#   classify(matrix(c(1, 0), ncol = 2, byrow = TRUE))
# 
# # Make sure that negative 1 km cells don't have any observed RTS in them outside of the negative bounding box
# mrts_neg_rast[mrts_neg_rast == 0 & !is.na(mrts_rast)] = NA
# 
# # Convert to data frame
# mrts_neg_grid = mrts_neg_rast |>
#   as.data.frame(xy = TRUE, na.rm = TRUE) |>
#   rename(count = last) |>
#   mutate(rts = count)
# 
# mrts_grid = mrts_grid |>
#   bind_rows(mrts_neg_grid)
# 
# rm(mrts_rast, mrts_neg_rast)
# 
# write_csv(mrts_grid, './arts_makopoulou_merge/mrts_grid.csv')

mrts_grid = read_csv('./arts_makopoulou_merge/mrts_grid.csv')
```

```{r}
# arts_grid = arts_grid |>
#   st_as_sf(coords = c('x', 'y'), crs = 3413)
# 
# mrts_grid = mrts_grid |>
#   st_as_sf(coords = c('x', 'y'), crs = 3413)
# 
# arts_mrts_merge = arts_grid |>
#   select(-count, arts = rts) |>
#   st_join(mrts_grid |>
#             select(-count, mrts = rts)) |>
#   bind_rows(
#     mrts_grid |>
#       select(-count, mrts = rts) |>
#       st_join(arts_grid |>
#                 select(-count, arts = rts))
#   ) |>
#   distinct() |>
#   mutate(
#     disagree = case_when(
#       arts == 1 & mrts == 0 ~ TRUE,
#       arts == 0 & mrts == 1 ~ TRUE,
#       TRUE ~ FALSE
#     ),
#     dataset = factor(
#       case_when(
#         !is.na(arts) & !is.na(mrts) ~ 'Both',
#         !is.na(arts) & is.na(mrts) ~ 'ARTS',
#         is.na(arts) & !is.na(mrts) ~ 'MRTS',
#         TRUE ~ NA
#       ),
#       levels = c('Both', 'ARTS', 'MRTS')
#     ),
#     class = factor(
#       case_when(
#         !(is.na(arts) & is.na(mrts)) & (arts == 1 | is.na(arts)) & (mrts == 1 | is.na(mrts)) ~ '+',
#         !(is.na(arts) & is.na(mrts)) & (arts == 0 | is.na(arts)) & (mrts == 0 | is.na(mrts)) ~ '-',
#         arts == 1 & mrts == 0 ~ '+/-',
#         arts == 0 & mrts == 1 ~ '-/+',
#         TRUE ~ NA
#       ),
#       levels = c('+', '-', '+/-', '-/+')
#     ),
#     dataset_class = factor(
#       case_when(
#         arts == 1 & mrts == 0 ~ 'Both - ARTS +',
#         arts == 0 & mrts == 1 ~ 'Both - MRTS +',
#         arts == 1 & mrts == 1 ~ 'Both +',
#         arts == 0 & mrts == 0 ~ 'Both -',
#         arts == 1 & is.na(mrts) ~ 'ARTS +',
#         arts == 0 & is.na(mrts) ~ 'ARTS -',
#         is.na(arts) & mrts == 1 ~ 'MRTS +',
#         is.na(arts) & mrts == 0 ~ 'MRTS -',
#         TRUE ~ NA
#       ),
#       levels = c(
#         'Both +',
#         'ARTS +',
#         'MRTS +',
#         'Both -',
#         'ARTS -',
#         'MRTS -',
#         'Both - ARTS +',
#         'Both - MRTS +'
#       )
#     )
#   )
# 
# st_write(arts_mrts_merge,
#          './arts_makopoulou_merge/arts_mrts_merge.geojson')

arts_mrts_merge = st_read('./arts_makopoulou_merge/arts_mrts_merge.geojson')
arts_mrts_merge_rast = rasterize(
  arts_mrts_merge,
  pca,
  field = 'dataset_class'
)

ggplot() + 
  geom_sf(data = world) +
  geom_sf(
    data = arts_mrts_merge |> filter(disagree == TRUE),
    aes(color = factor(arts))
  )
```


## Map Difference

```{r, fig.height = 6, fig.width = 6}
ggplot() +
  geom_sf(data = world) +
  geom_spatraster(
    data = rts_prob_diff,
    maxcell = 1e+07
  ) +
  scale_fill_gradient2(
    name = expression(Delta~' RTS Probability'),
    limits = c(-100, 100),
    low = '#0066CC',
    mid = '#FFFFDD',
    high = '#990000',
    na.value = 'transparent'
  ) +
  coord_sf(expand = FALSE) +
  ggtitle('RTS Probability Difference (Makopoulou - Woodwell)') +
  theme(legend.position = 'inside',
        legend.justification = c(1, 0),
        legend.position.inside = c(0.999, 0.001))
# ggsave('./figures/rts_prob_diff.jpg',
#        height = 6,
#        width = 6)

ggplot() +
  geom_sf(data = world) +
  geom_spatraster(
    data = rts_susceptibility_diff,
    maxcell = 1e+07
  ) +
  scale_fill_gradient2(
    name = expression(Delta~'RTS Susceptibility'),
    limits = c(-5, 5),
    low = '#0066CC',
    mid = '#FFFFDD',
    high = '#990000',
    na.value = 'transparent'
  ) +
  coord_sf(expand = FALSE) +
  ggtitle('RTS Susceptibility Difference (Makopoulou - Woodwell)')
```

```{r}
ggplot() +
  geom_sf(data = world) +
  geom_spatraster(
    data = rts_prob_diff,
    # maxcell = 1e+07
  ) +
  scale_fill_gradient2(
    name = expression(Delta~' RTS Probability'),
    limits = c(-100, 100),
    low = '#0066CC',
    mid = '#FFFFDD',
    high = '#990000',
    na.value = 'transparent'
  ) +
  geom_sf(
    data = arts_mrts_merge, 
    aes(color = class, shape = dataset), 
    size = 1,
    linewidth = 0.05
    ) +
  scale_shape_manual(
    name = 'RTS Source',
    values = c(16, 1, 6)
  ) +
  scale_color_manual(
    name = 'Training Class',
    values = c('black', 'white', 'gray30', 'gray70')
  ) +
  # scale_color_manual(
  #   name = 'RTS Source',
  #   values = c('#FFCC00', '#FFFF00', '#FFFF99', 
  #              '#660099', '#CC66CC', '#CCAACC', 
  #              '#009966', '#33FFCC')
  #   ) +
  # scale_shape_manual(
  #   name = 'RTS Source',
  #   values = c(
  #     16, 15, 18,
  #     1, 0, 5,
  #     7, 9
  #   )
  # ) +
  coord_sf(expand = FALSE) +
  ggtitle('RTS Probability Difference (Makopoulou - Woodwell)')

# ggsave('./figures/rts_susceptibility_diff.jpg',
#        height = 6,
#        width = 8)
```

## Differences Due to Training Data Differences

```{r}
arts_mrts_data_diff = c(arts_mrts_merge_rast, rts_prob_diff) |>
  as.data.frame(xy = TRUE, na.rm = TRUE) |>
  mutate(
    dataset_class = factor(
      case_when(
        dataset_class == 1 ~ levels(arts_mrts_merge$dataset_class)[1],
        dataset_class == 2 ~ levels(arts_mrts_merge$dataset_class)[2],
        dataset_class == 3 ~ levels(arts_mrts_merge$dataset_class)[3],
        dataset_class == 4 ~ levels(arts_mrts_merge$dataset_class)[4],
        dataset_class == 5 ~ levels(arts_mrts_merge$dataset_class)[5],
        dataset_class == 6 ~ levels(arts_mrts_merge$dataset_class)[6],
      ), 
      levels = levels(arts_mrts_merge$dataset_class)
    )
  ) |>
  rename('rts_prob_diff' = 'gbm_1_extr')
```

```{r}
ggplot(arts_mrts_data_diff, aes(x = dataset_class, y = rts_prob_diff)) +
  geom_hline(yintercept = 0, color = 'gray30', linetype = 'dashed') +
  geom_boxplot()
```

## Extract Underlying Variables for Principal Components

```{r}
components = pca_model$rotation |>
  as_tibble() |>
  mutate(
    variable = rownames(pca_model$rotation),
    .before = PC1
    ) |>
  select(variable:PC6) |>
  pivot_longer(PC1:PC6, names_to = 'pc', values_to = 'correlation') |>
  arrange(pc, -abs(correlation))

sd_factor = 1.5
main_components = components |>
  filter(correlation <= -1*sd_factor*sd(correlation) | correlation >= sd_factor*sd(correlation),
         .by = pc) |>
  mutate(order = seq(1:n()),
         label = paste0(variable, ': ', round(correlation, 2)),
         .by = pc) # |>
  # pivot_wider(
  #   id_cols = order, 
  #   names_from = PC, 
  #   values_from = c(variable, correlation),
  #   names_vary = 'slowest',
  #   names_glue = '{PC}_{.value}'
  #   )

# write_csv(
#   components,
#   './tables/pc_variable_correlations.csv'
# )
# write_csv(
#   main_components,
#   './tables/pc_main_components.csv'
# )
```

```{r}
ggplot(components,
       aes(x = correlation)) +
  geom_vline(xintercept = 0,
             color = 'gray50',
             linetype = 'dashed') +
  geom_histogram(bins = 30) +
  geom_histogram(
    data = main_components,
    aes(x = correlation),
    bins = 30,
    inherit.aes = FALSE,
    color = 'transparent',
    fill = 'black'
  ) +
  # geom_text(
  #   data = main_components,
  #   aes(x = correlation, y = 2.5, label = variable),
  #   angle = 90,
  #   hjust = 0,
  #   inherit.aes = FALSE
  #   ) +
  facet_wrap(~ pc, ncol = 2)
```

## Model PC Scores Relative to RTS Probability Difference

```{r}
rts_prob_diff_df = rts_prob_diff |>
  as.data.frame(
  xy = TRUE
) |>
  filter(!is.na(gbm_1_extr)) |>
  rename(prob_diff = gbm_1_extr) |>
  mutate(
    across(x:y,
           ~ round(.x, 8)),
    prob_diff_round = round(prob_diff/10, 0)*10,
    prob_diff_class = factor(prob_diff_round)
    )

pca_df = pca[[1:6]] |>
  mask(rts_prob_diff) |>
  as.data.frame(
    xy = TRUE
  ) |>
  filter(!is.na(PC1)) |>
  mutate(
    across(x:y,
           ~ round(.x, 8))
    )

rts_pca_df = rts_prob_diff_df |>
  left_join(pca_df,
            by = c('x', 'y')) |>
  pivot_longer(PC1:PC6, names_to = 'pc', values_to = 'value')
```

```{r}
rts_pca_df_summary = rts_pca_df |>
  summarize(
    value_mean = mean(value, na.rm = TRUE),
    value_median = median(value, na.rm = TRUE),
    value_sd = sd(value, na.rm = TRUE),
    .by = c(prob_diff_round, prob_diff_class, pc)
  )
```

```{r}
ggplot(
  rts_pca_df,
       aes(x = pc, 
           y = value, 
           color = prob_diff_class, 
           group = interaction(prob_diff_class, pc))
  ) +
  geom_boxplot(position = position_dodge()) +
  scale_y_continuous(name = 'PC Score') +
  scale_color_viridis(name = 'RTS\nProbability\nDifference',
                      discrete = TRUE) +
  theme(axis.title.x = element_blank())
```

```{r}
min_lm = lm(
  value_mean ~ 1,
  data = rts_pca_df_summary
)

max_lm = lm(
  value_mean ~ (prob_diff_round + I(prob_diff_round^2) + I(prob_diff_round^3))*pc,
  data = rts_pca_df_summary
)

rts_suscep_diff_summary_lm = step(
  max_lm,
  list(lower = min_lm, upper = max_lm)
)
summary(rts_suscep_diff_summary_lm)

rts_suscep_diff_lm_params = emmeans::emmeans(
  rts_suscep_diff_summary_lm,
  ~ pc,
  at = list(prob_diff_round = 0)
) |>
  as_tibble() |>
  rename(intercept = emmean) |>
  rename_with(~ paste0('int_', .x),
              .cols = SE:upper.CL) |>
  full_join(
    emmeans::emtrends(
      rts_suscep_diff_summary_lm, 
      ~ pc, 
      var = "prob_diff_round"
    ) |>
      as_tibble() |>
      rename(slope = prob_diff_round.trend) |>
      rename_with(~ paste0('slope_', .x),
                  .cols = SE:upper.CL)
  ) |>
  full_join(
    emmeans::emtrends(
      rts_suscep_diff_summary_lm, 
      ~ pc, 
      var = "I(prob_diff_round^2)"
    ) |>
      as_tibble() |>
      rename(quadratic_term = `I(prob_diff_round^2).trend`) |>
      rename_with(~ paste0('quadratic_', .x),
                  .cols = SE:upper.CL)
  ) |>
  full_join(
    emmeans::emtrends(
      rts_suscep_diff_summary_lm, 
      ~ pc, 
      var = "I(prob_diff_round^3)"
    ) |>
      as_tibble() |>
      rename(cubic_term = `I(prob_diff_round^3).trend`) |>
      rename_with(~ paste0('cubic_', .x),
                  .cols = SE:upper.CL)
  )
```

The Makopoulou model tends to predict higher RTS Susceptibility with low PC1, PC3, and PC4 values and high PC2 and PC5 values. In terms of input variables this means that the Makopoulou model has higher RTS Susceptibility when:
- precip sum, max, MAGT, runoff sum, and ALT are larger (PC1) - Marine influenced areas
- temp seasonality, tmax, srad sum, and LST are lower; and tmin and runoff max are higher (PC2) - Marine influenced areas
- the trend in soil moisture and LST over time are lower (PC5) - Less climate change related increase in precip/temperature
- the trend in precip, precip max, and runoff over time are higher; and silt values are higher (PC3)
- the trend in runoff and tmax over time are higher; silt and MAGT are lower; and soil moisture and said are higher

```{r, fig.height = 6, fig.width = 6.5}
labels = tibble(
      pc = paste0('PC', c(seq(1, 6), 6)),
      x = c(rep(NA, 5), -5, 5),
      hjust = c(rep(NA, 5), 1, 0),
      xend = c(rep(NA, 5), -40, 40),
      y = c(rep(NA, 5), -8, -8),
      label = c(rep(NA, 5), 'Woodwell Greater', 'Makopoulou Greater'),
      yarrow = c(rep(NA, 5), -7, -7),
      )

ggplot(
  rts_pca_df_summary,
       aes(x = prob_diff_round, 
           y = value_mean)
  ) +
  geom_vline(xintercept = 0,
             color = 'gray50',
             linetype = 'dashed') +
  # geom_point(
  #   aes(y = value_median),
  #   size = 2,
  #   color = 'red'
  #   ) +
  geom_point(size = 0.5) +
  geom_errorbar(
    aes(ymin = value_mean - value_sd, ymax = value_mean + value_sd)
    ) +
  geom_smooth(
    method = 'lm', 
    formula = y ~ (x + I(x^2) + I(x^3)),
    color = 'black',
    linewidth = 0.5
    ) +
  # geom_smooth(
  #   method = 'gam', 
  #   linewidth = 0.5
  #   ) +
  geom_text(
    data = labels,
    aes(x = x, y = y, label = label, hjust = hjust),
    size = 2.5
  ) +
  geom_segment(
    data = labels,
    aes(x = x, xend = xend, y = yarrow, yend = yarrow),
    arrow = arrow(length = unit(0.05, 'inches'))
    ) +
  geom_text(
    data = main_components,
    aes(x = -100, y = -3.5 - order*0.75, label = label),
    hjust = 0,
    vjust = 0,
    size = 2.5
  ) +
  scale_x_continuous(
    name = 'RTS Probability Difference\n(Makopoulou-Woodwell)',
    # breaks = seq(-4, 4, by = 2),
    ) +
  scale_y_continuous(name = 'PC Score') +
  facet_wrap(~ pc, ncol = 3) +
  ggtitle('PC Scores by Mean RTS Probability Difference\nWith Cubic Linear Models')

# ggsave('./figures/rts_prob_diff_models.jpg',
#        height = 6,
#        width = 6.5)
```

## Model Environmental Variables Relative to RTS Probability Difference

```{r}
m_env_df = m_env[[1:6]] |>
  project(rts_prob_diff) |>
  mask(rts_prob_diff) |>
  as.data.frame(
    xy = TRUE
  ) |>
  filter(!is.na(bd) & !is.na(fdd) & !is.na(fine) & !is.na(rainfall) & !is.na(slop90_geom_A) & !is.na(tdd)) |>
  mutate(
    across(x:y,
           ~ round(.x, 8)),
    across(bd:tdd, 
           ~ (.x - mean(.x, na.rm = TRUE))/sd(.x))
    )

rts_m_env_df = rts_prob_diff_df |>
  left_join(m_env_df,
            by = c('x', 'y')) |>
  pivot_longer(bd:tdd, names_to = 'variable', values_to = 'value')
```

```{r}
rts_m_env_df_summary = rts_m_env_df |>
  summarize(
    value_mean = mean(value, na.rm = TRUE),
    value_median = median(value, na.rm = TRUE),
    value_sd = sd(value, na.rm = TRUE),
    .by = c(prob_diff_round, prob_diff_class, variable)
  )
```

```{r}
ggplot(
  rts_m_env_df_summary,
       aes(x = prob_diff_round, 
           y = value_mean)
  ) +
  geom_vline(xintercept = 0,
             color = 'gray50',
             linetype = 'dashed') +
  # geom_point(
  #   aes(y = value_median),
  #   size = 2,
  #   color = 'red'
  #   ) +
  geom_point(size = 0.5) +
  geom_errorbar(
    aes(ymin = value_mean - value_sd, ymax = value_mean + value_sd)
    ) +
  geom_smooth(
    method = 'lm', 
    formula = y ~ (x + I(x^2) + I(x^3)),
    color = 'black',
    linewidth = 0.5
    ) +
  # geom_text(
  #   data = labels,
  #   aes(x = x, y = y, label = label, hjust = hjust),
  #   size = 2.5
  # ) +
  # geom_segment(
  #   data = labels,
  #   aes(x = x, xend = xend, y = yarrow, yend = yarrow),
  #   arrow = arrow(length = unit(0.05, 'inches'))
  #   ) +
  # geom_text(
  #   data = main_components,
  #   aes(x = -100, y = -3.5 - order*0.75, label = label),
  #   hjust = 0,
  #   vjust = 0,
  #   size = 2.5
  # ) +
  scale_x_continuous(
    name = 'RTS Probability Difference\n(Makopoulou-Woodwell)',
    # breaks = seq(-4, 4, by = 2),
    ) +
  scale_y_continuous(name = 'Environmental Variable Z-Score') +
  facet_wrap(~ variable, ncol = 3) +
  ggtitle('Environmental Variable Scores by Mean RTS Probability Difference\nWith Cubic Linear Models')

# ggsave('./figures/rts_prob_diff_models_makopoulou.jpg',
#        height = 6,
#        width = 6.5)

```

## Map Top Predictors of RTS in Our Model

```{r}
ggplot() +
  geom_sf(data = world) +
  geom_spatraster_rgb(
    data = pca,
    r = 2,
    g = 1,
    b = 5,
    max_col_value = 7
  ) +
  # geom_sf(data = rts_susceptibility_diff_extremes_poly,
  #         aes(color = susceptibility_class),
  #         fill = 'transparent') +
  coord_sf(expand = FALSE) +
  ggtitle('Principal Components 2, 1, and 5 as RGB')

# ggsave('./figures/top_predictors_rgb.jpg',
#        height = 6.5,
#        width = 6.5)
```

