---
title: "RTS Representation"
author: "Heidi Rodenhizer"
date: "`r Sys.Date()`"
output: html_document
---

# Load Packages

```{r}
library(terra)
library(tidyterra)
library(sf)
library(viridis)
library(bestNormalize)
library(ggnewscale)
library(tidyverse)
```

```{r}
theme_set(theme_bw())
```


# Functions

## calc_data_need()

```{r}
calc_data_need = function(prob_rast, dist_rast, new_name) {
  data_need = prob_rast - dist_rast
  data_need[data_need < 0] = 0
  names(data_need) = c(new_name)

  return(data_need)
}
```

## normalize_raster()

```{r}
normalize_raster = function(input_rast) {
  rast_rescale = input_rast

  values(rast_rescale) = predict(
    orderNorm(values(rast_rescale))
  )

  rast_rescale = (rast_rescale -
    as.numeric(
      global(rast_rescale, fun = 'min', na.rm = TRUE)
    ))

  rast_rescale = rast_rescale /
    as.numeric(
      global(
        rast_rescale,
        fun = 'max',
        na.rm = TRUE
      )
    )

  names(rast_rescale) = paste0(names(rast_rescale), "_rescale")

  return(rast_rescale)
}
```

## rename_inline

```{r}
rename_inline = function(obj, name_vec) {
  names(obj) = name_vec
  return(obj)
}
```


# Load Data

```{r}
rts_prob = list(
  rast('./sdm/gbm_kfold_10_pred_tuned_median_v.1.0.1_filter.tif') / 100,
  rast('./sdm/gbm_kfold_10_pred_tuned_median_v.3.1.0_filter.tif') / 100
) |>
  map(
    ~ .x |>
      rename_inline(c("rts_prob"))
  )
names(rts_prob) = c("rts_prob_v1", "rts_prob_v3")

rts_prob_m = rast(
  '../makopoulou_data/makopoulou_data_download_20240916/GBM/gbm_aver_raster_calc_10_runs.tif'
) |>
  project(rts_prob[["rts_prob_v1"]], method = 'near')
rts_prob_m = rts_prob_m / 1000
names(rts_prob_m) = c("rts_prob")

euc_dist = list(
  rast('./euclidean_distance/polygon_counts_1_2_v.1.0.1.tif'),
  rast('./euclidean_distance/polygon_counts_1_2_v.3.1.0.tif')
)
names(euc_dist) = c("euc_dist_v1", "euc_dist_v3")

rts_train_new = st_read(
  "./sdm/training_data/rts_train_new_v.3.1.0-v.1.0.1.geojson"
)
```

```{r}
world = rnaturalearth::ne_countries(scale = "large", returnclass = "sf")
world = world |>
  st_transform(3413) |>
  st_crop(ext(rts_prob[["rts_prob_v1"]]))
```

# Prep Hex Layer of New RTS

```{r}
rts_train_centroid = rts_train_new %>%
  st_centroid() %>%
  bind_cols(st_coordinates(.))
```


# Histograms

## Euclidean Distance

What normalization method is best?

```{r}
euc_dist_df = map(
  euc_dist,
  ~ .x |>
    subset('dist_2_count') |>
    as.data.frame(xy = TRUE, na.rm = FALSE)
)

# bestNormalize(
#   euc_dist_df |>
#     slice_sample(prop = 0.1) |>
#     pull(dist_2_count)
# )
```

Normalize

```{r}
euc_dist_rescale = euc_dist |>
  map(
    ~ .x |>
      select("dist_2_count") |>
      normalize_raster()
  )

euc_dist_df = euc_dist_df |>
  map2(
    euc_dist_rescale,
    ~ .x |>
      full_join(
        .y |>
          as.data.frame(xy = TRUE, na.rm = FALSE),
        by = c("x", "y")
      )
  )
```

Visualize

```{r}
max_dist_2_count = euc_dist_df |>
  map(
    ~ .x |>
      pull(dist_2_count) |>
      max(na.rm = TRUE)
  )

map2(
  euc_dist_df,
  max_dist_2_count,
  ~ ggplot(.x) +
    geom_histogram(
      aes(x = dist_2_count / max_dist_2_count[[1]]),
      color = "darkred",
      fill = "darkred",
      alpha = 0.5
    ) +
    geom_histogram(
      aes(x = dist_2_count_rescale),
      color = "cadetblue1",
      fill = "cadetblue1",
      alpha = 0.5
    )
)
```

## RTS Probability (Woodwell)

What normalization method is best?

```{r}
rts_prob_df = rts_prob |>
  map(
    ~ .x |>
      as.data.frame(xy = TRUE, na.rm = FALSE)
  )

# bestNormalize(
#   rts_prob_df |>
#     slice_sample(prop = 0.1) |>
#     pull(median)
# )
```

Normalize

```{r}
rts_prob_rescale = rts_prob |>
  map(
    ~ .x |>
      normalize_raster()
  )

rts_prob_df = rts_prob_df |>
  map2(
    rts_prob_rescale,
    ~ .x |>
      full_join(
        .y |>
          as.data.frame(xy = TRUE, na.rm = FALSE),
        by = c("x", "y")
      )
  )
```

Visualize

```{r}
map(
  rts_prob_df,
  ~ ggplot(.x) +
    geom_histogram(
      aes(x = rts_prob),
      color = "darkred",
      fill = "darkred",
      alpha = 0.5
    ) +
    geom_histogram(
      aes(x = rts_prob_rescale),
      color = "cadetblue1",
      fill = "cadetblue1",
      alpha = 0.5
    )
)
```

## RTS Probability (Makopoulou)

What normalization method is best?

```{r}
rts_prob_m_df = rts_prob_m |>
  as.data.frame(xy = TRUE, na.rm = FALSE)

# bestNormalize(
#   rts_prob_m_df |>
#     slice_sample(prop = 0.1) |>
#     pull(gbm_1_extr)
# )
```

Normalize

```{r}
rts_prob_m_rescale = rts_prob_m |>
  normalize_raster()

rts_prob_m_df = rts_prob_m_df |>
  full_join(
    rts_prob_m_rescale |>
      as.data.frame(xy = TRUE, na.rm = FALSE),
    by = c("x", "y")
  )
```

Visualize

```{r}
ggplot(rts_prob_m_df) +
  geom_histogram(
    aes(x = rts_prob),
    color = "darkred",
    fill = "darkred",
    alpha = 0.5
  ) +
  geom_histogram(
    aes(x = rts_prob_rescale),
    color = "cadetblue1",
    fill = "cadetblue1",
    alpha = 0.5
  )
```


# Weight RTS Representation

## Woodwell

```{r}
# rts_rep_v1 = c(
#   calc_data_need(
#     rts_prob_rescale[["rts_prob_v1"]],
#     euc_dist_rescale[["euc_dist_v1"]],
#     "data_need_rescale_both"
#   ),
#   calc_data_need(
#     rts_prob[["rts_prob_v1"]],
#     euc_dist_rescale[["euc_dist_v1"]],
#     "data_need_rescale_dist"
#   )
# )

# writeRaster(
#   rts_rep_v1,
#   './data_gaps/rts_data_gaps_filter_v.1.0.1.tif'
#   )
rts_rep_v1 = rast('./data_gaps/rts_data_gaps_filter_v.1.0.1.tif')

# rts_rep_v3 = c(
#   calc_data_need(
#     rts_prob_rescale[["rts_prob_v3"]],
#     euc_dist_rescale[["euc_dist_v3"]],
#     "data_need_rescale_both"
#   ),
#   calc_data_need(
#     rts_prob[["rts_prob_v3"]],
#     euc_dist_rescale[["euc_dist_v3"]],
#     "data_need_rescale_dist"
#   )
# )

# writeRaster(
#   rts_rep_v3,
#   './data_gaps/rts_data_gaps_filter_v.3.1.0.tif'
#   )
rts_rep_v3 = rast('./data_gaps/rts_data_gaps_filter_v.3.1.0.tif')

rts_rep_improvement = rts_rep_v3 - rts_rep_v1
```

```{r}
ggplot() +
  geom_sf(data = world, fill = "gray95") +
  geom_spatraster(
    data = rts_rep_v1 |>
      subset("data_need_rescale_both"),
    maxcell = 1e+07
  ) +
  scale_fill_viridis(
    name = 'Data Need\n[0,1]',
    # limits = c(0, 1),
    na.value = 'transparent'
  ) +
  geom_sf(data = world, fill = "transparent") +
  coord_sf(expand = FALSE)

ggsave(
  './figures/rts_data_need_woodwell_filter_v.1.0.1.jpg',
  height = 6,
  width = 8
)

ggplot() +
  geom_sf(data = world, fill = "gray95") +
  geom_spatraster(
    data = rts_rep_v3 |>
      subset("data_need_rescale_both"),
    maxcell = 1e+07
  ) +
  scale_fill_viridis(
    name = 'Data Need\n[0,1]',
    # limits = c(0, 1),
    na.value = 'transparent'
  ) +
  geom_sf(data = world, fill = "transparent") +
  coord_sf(expand = FALSE)

# ggsave(
#   './figures/rts_data_need_woodwell_filter_v.3.1.0.jpg',
#   height = 6,
#   width = 8
# )
```

```{r}
ggplot() +
  geom_sf(data = world, fill = "gray95") +
  geom_spatraster(
    data = rts_rep_improvement |>
      subset("data_need_rescale_both"),
    # maxcell = 1e+07
  ) +
  scale_fill_gradient2(
    name = expression(atop(Delta ~ " Data Need", "(v3-v1)")),
    limits = c(-0.5, 0.5),
    low = '#0066CC',
    mid = '#FFFFDD',
    high = '#990000',
    na.value = 'transparent',
    oob = scales::squish
  ) +
  geom_sf(data = world, fill = "transparent") +
  coord_sf(expand = FALSE) +
  theme(axis.title = element_blank())

# ggsave(
#   './figures/rts_data_need_improvement.jpg',
#   height = 6,
#   width = 8
# )

ggplot() +
  geom_sf(data = world, fill = "gray95") +
  geom_spatraster(
    data = rts_rep_improvement |>
      subset("data_need_rescale_both"),
    maxcell = 1e+07
  ) +
  scale_fill_gradient2(
    name = expression(atop(Delta ~ " Data Need", "(v3-v1; [-1,1])")),
    limits = c(-0.25, 0.25),
    low = '#0066CC',
    mid = '#FFFFDD',
    high = '#990000',
    na.value = 'transparent',
    oob = scales::squish,
    guide = guide_colorbar(order = 1)
  ) +
  # new_scale_fill() +
  geom_hex(
    data = rts_train_centroid |>
      filter(TrainClass == "Positive") |>
      mutate(count = 1),
    aes(X, Y, color = after_stat(count)),
    fill = "transparent"
  ) +
  scale_color_continuous(
    name = "Added RTS",
    guide = guide_legend(order = 2),
    trans = "log10"
  ) +
  geom_sf(data = world, fill = "transparent") +
  coord_sf(expand = FALSE) +
  theme(axis.title = element_blank())

# ggsave(
#   './figures/rts_data_need_improvement.jpg',
#   height = 6,
#   width = 8
# )
```

## Makopoulou

```{r}
rts_rep_m = euc_dist_rescale * rts_probability_rescale_m
rts_rep_m = rts_rep_weighted_m / minmax(rts_rep_weighted_m)[2]
names(rts_rep_weighted_m) = c('rts_data_need')

# writeRaster(
#   rts_rep_weighted_m,
#   './data_gaps/rts_data_gaps_makopoulou.tif'
#   )
```

```{r}
ggplot() +
  geom_sf(data = world) +
  geom_spatraster(data = rts_rep_weighted_m, maxcell = 1e+07) +
  scale_fill_viridis(name = 'Need for Data', na.value = 'transparent') +
  coord_sf(expand = FALSE) +
  ggtitle('RTS Knowledge Gaps Weighted by RTS Probability (Makopoulou)')

# ggsave('./figures/rts_data_need_makopoulou.jpg',
#        height = 6,
#        width = 8)
```

```{r}
ggplot(
  rts_rep_weighted |> as.data.frame(xy = TRUE, na.rm = TRUE),
  aes(x = rts_data_need)
) +
  geom_histogram()

ggplot(
  rts_rep_weighted_m |> as.data.frame(xy = TRUE, na.rm = TRUE),
  aes(x = rts_data_need)
) +
  geom_histogram()
```

```{r}
ggplot() +
  geom_sf(data = world) +
  geom_spatraster(data = rts_rep_weighted > 0.3, maxcell = 1e+07) +
  scale_fill_viridis(
    name = 'Need for Data',
    discrete = TRUE,
    na.value = 'transparent',
    labels = c('High', 'Low'),
    na.translate = FALSE
  ) +
  coord_sf(expand = FALSE) +
  ggtitle('RTS Knowledge Gaps Weighted by RTS Probability (Woodwell)')

ggplot() +
  geom_sf(data = world) +
  geom_spatraster(data = rts_rep_weighted_m > 0.3, maxcell = 1e+07) +
  scale_fill_viridis(
    name = 'Need for Data',
    discrete = TRUE,
    na.value = 'transparent',
    labels = c('High', 'Low'),
    na.translate = FALSE
  ) +
  coord_sf(expand = FALSE) +
  ggtitle('RTS Knowledge Gaps Weighted by RTS Probability (Makopoulou)')

```

