---
title: "Euclidean Distance"
author: "Heidi Rodenhizer"
date: "`r Sys.Date()`"
output: html_document
---

# Load Libraries

```{r}
library(rnaturalearth)
library(svMisc)
library(terra)
library(tidyterra)
library(sf)
library(foreach)
library(doParallel)
library(viridis)
library(tidyverse)
```

```{r}
theme_set(theme_bw())
```


# Load Data

```{r}
km = rast('./kmeans/kmeans_10.tif')
levels(km) = tibble(id = seq(1, 20),
                    kmeans = seq(1, 20))

pca = rast('./pca/pca.tif') |>
  subset(c('PC1', 'PC2', 'PC3', 'PC4', 'PC5', 'PC6'))

pca_df = pca |>
  as.data.frame(cell = TRUE, xy = TRUE, na.rm = TRUE)

arts_pc = st_read('./pca/arts_pc_v.3.1.0.geojson') |>
  st_drop_geometry() |>
  select(-matches('PC([789]{1}|\\d{2})'))|> # only keep PC1-PC6
  filter(
    !(is.na(PC1) | is.na(PC2) | is.na(PC3) | is.na(PC4) | is.na(PC5) | is.na(PC6))
    ) 

world <- ne_countries(scale = "large", returnclass = "sf")

arts_pc_plot = st_read('./pca/arts_pc_v.3.1.0.geojson') |>
  filter(
    !(is.na(PC1) | is.na(PC2) | is.na(PC3) | is.na(PC4) | is.na(PC5) | is.na(PC6))
    )
```


# Euclidean Distance

## Polygon Count Within X Distance

```{r}
# converting to list prior to for loop speeds up computation
# ENORMOUSLY
arts_pc_l = arts_pc |>
  select(PC1:PC6) |>
  drop_na() |>
  as.matrix() |>
  t() |>
  as_tibble() |>
  as.list()
```

```{r}
# converting to list prior to for loop speeds up computation
# ENORMOUSLY
# As odd as it may seem, it is faster to convert to matrix, transpose, convert back to tibble and then convert to list (because a tibble's columns are already lists) than to group_split() the dataframe
# sample_size = 10000

pca_l = pca_df |>
  # slice(sample(nrow(pca_df), size = sample_size)) |>
  select(PC1:PC6) |>
  as.matrix() |>
  t() |>
  as_tibble() |>
  as.list()
```

Try cut-offs of 2 and 1.

2 seems pretty close, and is about the standard deviation of euclidean distances between each pixel and all of the RTS polygons.

1 - Calculate euclidean distance to each polygon from each pixel.
2 - Calculate number of polygons closer than 3, 2, and 1 unit distance.
3 - Save three values as bands in a raster.

This requires ~32 seconds for a subset of 10000 pixels, or about 12.3 hours for the whole dataset with the current number of polygons.

Running on 18 cores in parallel, this takes about 9.4 seconds for 10000 pixels, or about 3.6 hours for all pixels over ~2100 polygons.

For version v.1.0.1, with 39661 polygons with environmental data, with the first 6 PC layers and distances of 1 or 2, a subset of 10000 pixels took 2.29 minutes (138 seconds) to run. For the whole region, it should take about 53 hours, or 2.2 days. It took 1.44 days.

For version v.3.1.0, with 49046 polygons with environmental data, with the first 6 PC layers and distances of 1 or 2, we would expect it to take 1.78 days (based on the actual processing time from v.1.0.1). It took 1.97 days.

```{r}
n_polys = length(arts_pc_l)
n_pixels = length(pca_l)

time1 = Sys.time()

counts = matrix(
  nrow = n_pixels,
  ncol = 3
)

cl = makeCluster(18) # I have 20 cores, this leaves me with 2 to do other stuff
registerDoParallel(cl)

counts = foreach (pixel_idx = 1:n_pixels) %dopar% {

  euc_dist = vector(mode = 'double', length = n_polys)

  for (poly_idx in 1:n_polys) {

    euc_dist[poly_idx] = philentropy:: euclidean(
      pca_l[[pixel_idx]],
      arts_pc_l[[poly_idx]],
      testNA = FALSE
    )

  }

  c(
      length(euc_dist[euc_dist <= 1]),
      length(euc_dist[euc_dist <= 2]),
      mean(euc_dist)
    )

}

time2 = Sys.time()

print(time2 - time1)

stopCluster(cl)

counts = matrix(
  unlist(counts),
  byrow = TRUE,
  nrow = length(counts)
  )
```

```{r}
euc_dist_rast = pca[[1:3]]
values(euc_dist_rast[[1]])[pca_df$cell] = counts[,1]
values(euc_dist_rast[[2]])[pca_df$cell] = counts[,2]
values(euc_dist_rast[[3]])[pca_df$cell] = counts[,3]

names(euc_dist_rast) = c('dist_1_count', 'dist_2_count', 'mean_dist')

writeRaster(euc_dist_rast, './euclidean_distance/polygon_counts_1_2_v.3.1.0.tif')
```

```{r}
euc_dist_rast = rast('./euclidean_distance/polygon_counts_1_2_v.3.1.0.tif')
world = world |>
  st_transform(3413) |>
  st_crop(ext(euc_dist_rast))
```

```{r}
ggplot() +
  geom_sf(data = world, 
          aes(geometry = geometry),
          inherit.aes = FALSE,
          fill = 'gray95') +
  geom_spatraster(data = euc_dist_rast |>
                    subset(1:3),
                  maxcell = 1e+07) +
  geom_sf(data = world, 
          aes(geometry = geometry),
          inherit.aes = FALSE,
          fill = 'transparent') +
  scale_fill_viridis(name = 'Polygon Count',
                     na.value = 'transparent',
                     # limits = c(0, 100),
                     oob = scales::squish) +
  facet_wrap(~lyr, ncol = 2) +
  coord_sf(expand = FALSE) +
  theme(legend.position = 'inside',
        legend.position.inside = c(0.75, 0.25),
        legend.justification = c(0.5, 0.5))

# ggsave('./figures/euc_dist_poly_counts_v.3.1.0.jpg',
#        height = 6.5, width = 6.5)
# ggsave('./figures/euc_dist_poly_counts_v.3.1.0.pdf',
#        height = 6.5, width = 6.5)

ggplot() +
  geom_sf(data = world, 
          aes(geometry = geometry),
          inherit.aes = FALSE,
          fill = 'gray95') +
  geom_spatraster(data = euc_dist_rast |>
                    subset(2),
                  maxcell = 1e+07) +
  geom_sf(data = world, 
          aes(geometry = geometry),
          inherit.aes = FALSE,
          fill = 'transparent') +
  scale_fill_viridis(name = 'Polygon Count\n(Dist = 2)',
                     na.value = 'transparent',
                     # limits = c(0, 15),
                     oob = scales::squish) +
  facet_wrap(~lyr, ncol = 2) +
  coord_sf(expand = FALSE) #+
  # theme(legend.position = 'inside',
  #       legend.position.inside = c(0.75, 0.25),
  #       legend.justification = c(0.5, 0.5))

# ggsave('./figures/euc_dist_poly_counts_2_v.3.1.0.jpg',
#        height = 6.5, width = 6.5)
# ggsave('./figures/euc_dist_poly_counts_2_v.3.1.0.pdf',
#        height = 6.5, width = 6.5)

ggplot() +
  geom_sf(data = world, 
          aes(geometry = geometry),
          inherit.aes = FALSE,
          fill = 'gray95') +
  geom_spatraster(data = euc_dist_rast |>
                    subset(3),
                  maxcell = 1e+07) +
  geom_sf(data = world, 
          aes(geometry = geometry),
          inherit.aes = FALSE,
          fill = 'transparent') +
  scale_fill_viridis(name = 'Mean\nEuclidean\nDistance',
                     na.value = 'transparent',
                     limits = c(0, 15),
                     oob = scales::squish) +
  facet_wrap(~lyr, ncol = 2) +
  coord_sf(expand = FALSE) #+
  # theme(legend.position = 'inside',
  #       legend.position.inside = c(0.75, 0.25),
  #       legend.justification = c(0.5, 0.5))

# ggsave('./figures/mean_euc_dist_v.3.1.0.jpg',
#        height = 6.5, width = 6.5)
# ggsave('./figures/mean_euc_dist_v.3.1.0.pdf',
#        height = 6.5, width = 6.5)
```

## Improvement

```{r}
euc_dist_rast_old = rast('./euclidean_distance/polygon_counts_1_2_v.1.0.1.tif')
```

```{r}
euc_dist_diff = euc_dist_rast - euc_dist_rast_old
```

```{r}
ggplot() +
  geom_sf(data = world, 
          aes(geometry = geometry),
          inherit.aes = FALSE,
          fill = 'gray95') +
  geom_spatraster(data = euc_dist_diff |>
                    subset(2),
                  maxcell = 1e+07) +
  scale_fill_viridis(name = 'Improvement in\nPolygon Count\n(Dist = 2)',
                     na.value = 'transparent',
                     # limits = c(0, 15),
                     oob = scales::squish) +
  facet_wrap(~lyr, ncol = 2) +
  coord_sf(expand = FALSE) #+
  # theme(legend.position = 'inside',
  #       legend.position.inside = c(0.75, 0.25),
  #       legend.justification = c(0.5, 0.5))

# ggsave('./figures/euc_dist_poly_counts_2_improvement.jpg',
#        height = 6.5, width = 6.5)
# ggsave('./figures/euc_dist_poly_counts_2_improvement.pdf',
#        height = 6.5, width = 6.5)
```
